Using \LaTeX\ to submit manuscripts to peer reviewed journals.

### 2014-02-21 docs/thesis/appendix-conf_files.tex
work on document: added graphics

\begin{verbatim}
   <!-- Arthur Oviedo AO-->
   <!-- Fields declaration using the field types declared below-->
   <field name="math_notational_field" 
   		type="math_notational_type"/>
   <field name="math_structural_field" 
   		type="math_structural_type"/>
   <field name="math_normalized_notational_field" 
   		type="math_normalized_notational_type"/>
   <field name="math_full_normalized_notational_field" 
   		type="math_full_normalized_notational_type"/>
   <field name="filename" 
   		type="string" indexed="true" stored="true"/>
   <field name="number_occurences" 
   		type="int" indexed="true" stored="true"/>
\end{verbatim}

>>>>

\begin{verbatim}
   <!-- Arthur Oviedo AO-->
   <!-- Fields declaration using the field types declared below-->
   <field name="math_notational_field" 
   		type="math_notational_type"/>
   <field name="math_structural_field" 
   		type="math_structural_type"/>
   <field name="math_normalized_notational_field" 
   		type="math_normalized_notational_type"/>
   <field name="filename" 
   		type="string" indexed="true" stored="true"/>
   <field name="number_occurrences" 
   		type="int" indexed="true" stored="true"/>
\end{verbatim}

### 2014-02-21 docs/thesis/chap-cern_math_explorer.tex
work on document: added graphics

After examining a set of relevant documents from our dataset, we identified some heuristics that could be applied to improve the recall of our system:

@@

\subsection{Structural Features}

\subsubsection{Algebraic structures}
The first set of structures consist on purely algebraic relationships between the expressions.
Inspired by predefined integral tables for different kinds of expressions, we decided to extract from a given equation, whether it has the occurrence of these kind of patterns.
An example of this type of structures would be: $X\textunderscore*(X\textunderscore + A\textunderscore)$ where $X\textunderscore$ and $A\textunderscore$ can be any type of expression. Sample equations matching this pattern can be $C (C+1)$ or $\frac{(\sin{\theta}\cos{\theta})^2}{2a} (\frac{(\sin{\theta}\cos{\theta})^2}{2a} + \tan{\frac{\theta+\phi}{2}})$. This structural analysis of equations is more powerful than the unification of variables and constants proposed in previously discussed projects because these patterns, as shown in the second example, can match an arbitrary complex expression. The drawback of our solution is the need to manually select representative and useful patterns. This selection was done based on standard based tables of integrals similar to the ones that Computational Algebraic Systems employ to compute them.

@@

\subsection{Development with Lucene/Solr}
This section explains how the described features were implemented in the {\codefont Apache Solr/Lucene} framework.
\subsection{Integration with Mathematica}
This section describes some aspects that should be had to be considered while implementing the integration with the Mathematica CAS. Our system uses the provided {\codefont JLink} library to create a connection with the underlying process {\codefont MathKernel}. The main class for communicating java code and {\codefont Mathematica} is {\codefont KernelLink} and the main used method was {\codefont evaluateToOutputForm(String expr, int pageWidth)} where {\codefont expr} is the expression to be evaluated and {\codefont pageWidth} affects the formatting of the result.

@@

\subsection{Integration with CDS}

>>>>

After examining a set of documents from our dataset, we identified some heuristics that could be applied to improve the recall of our system:

@@

\subsection{Structural Features}
The structure of a mathematical equations is one of the most important features that can be extracted.

\subsubsection{Algebraic structures}
The first set of structures consist on purely algebraic relationships between the expressions.
Inspired by predefined integral tables for different kinds of expressions, we decided to extract from a given equation, whether it has the occurrence of these kind of patterns.
An example of this type of structures would be: $X\textunderscore*(X\textunderscore + A\textunderscore)$ where $X\textunderscore$ and $A\textunderscore$ can be any type of expression. Sample equations matching this pattern can be $C (C+1)$ or $\frac{(\sin{\theta}\cos{\theta})^2}{2a} (\frac{(\sin{\theta}\cos{\theta})^2}{2a} + \tan{\frac{\theta+\phi}{2}})$. This structural analysis of equations is more powerful than the unification of variables and constants proposed in previously discussed projects because these patterns, as shown in the second example, can match an arbitrary complex expression. The drawback of our solution is the need to manually select representative and useful patterns. This selection was done based on standard based tables of integrals similar to the ones that CASs employ to compute them.

@@

\subsection{Development with Lucene/Solr}
This section explains how the described components were implemented in the {\codefont Apache Solr/Lucene} framework.
One focus of our work was to develop a modular system that allowed easily to deploy new heuristics as they were identified.

Lucene framework defines some key concepts which ultimately are mapped into concrete classes. A {\codefont Document} is the minimum retrieval unit. For $5e^{x+y}$ we could either use a complete article as a document or a single equation. We choose the latter because it makes easier to focus on extracting features from a single equation. A document contains a set of {\codefont fields} which represent different types of information that it can store. Each field has different options such as whether it would be indexed and/or stored. A field can also indicate whether it will be a {\codefont Term Vector}. Our fields are as follow: 
\begin{itemize}
\item math\_notational\_field: This field stores the notational features as a Term Vector. 
\item math\_structural\_field: This field stores the structural features as a Term Vector. 
\item math\_normalized\_notational\_field: This field stores the notational features of the normalized expression as a Term Vector. 
\item filename: This is an auxiliary field which stores the CDS record / article name where the equation was found.
\item number\_occurrences: Another auxiliary field that is used to score final results.
\end{itemize}

Each field is associated to a particular {\codefont Analyzer}. An analyzer is a specific class that handles an element from a field and groups all the operations that are to be performed. For this purpose we wrote {\codefont SolrNotationalAnalyzer}, {\codefont SolrStructuralAnalyzer} and {\codefont SolrNormalizedNotationalAnalyzer}. 
The Solr prefix is added to make clear the fact that the class extends a {\codefont SolrAnalyzer} which extends a normal lucene {\codefont Analyzer}.

An analyzer consists of a sequence of {\codefont TokenStrem} elements. Normally an analyzer contains one {\codefont Tokenizer} element (which extends a {\codefont TokenStrem}) that is in charge of creating the Terms that will be stored in the field, and one or more {\codefont TokenFilter} elements which perform operations on the single tokens that were created previously. For the specific cases of SolrNotationalAnalyzer and SolrNormalizedNotationalAnalyzer we created the following classes:

\begin{itemize}
\item {\codefont MultiplePatternTokenizer}: This class is in charge of applying a set of regular expressions over the given expression in order to extract the single elements that were presented. Lucene incorporates the class {\codefont PatternTokenizer} but it only accepts one regular expression.
\item {\codefont UnicodeNormalizingFilter}: This class receives single tokens and verifies if the token is in Unicode NFKD. In negative case, it emits an additional token in the normalized form.
\item {\codefont SynonymExpandingFilter}: This class verifies if the received token belongs to a predefined category of elements and emits an additional token with it.
\item {\codefont NumericalRoundingFilter}: This class process {\codefont <mn>} tokens. When a number contains a decimal part, it applies the standard rounding mode iteratively and for each one, it emits a token for each step. 
\end{itemize}

{\codefont SolrStructuralAnalyzer } is simpler in its workflow: It consits of a {\codefont StructuralPatternsTokenizer} that tries to find occurrences of the predefined patterns in the expression. For each match, it emits a token with the pattern that was found. 
\subsection{Integration with Mathematica}
This section describes some aspects that had to be considered while implementing the integration with Mathematica. Our system uses the {\codefont JLink} library to create a connection with the underlying process {\codefont MathKernel}. The main class for communicating java code and {\codefont Mathematica} is {\codefont KernelLink} and the main used method was {\codefont evaluateToOutputForm(String expr, int pageWidth)} where {\codefont expr} is the expression to be evaluated and {\codefont pageWidth} affects the formatting of the result.

@@

\subsection{Integration with CDS}
Currently the Invenio code-base is being under a deep refactoring and rewriting. CDS currently runs in the stable branch {\codefont master} while the new developments are preferred to be implemented in the new branch {\codefont pu}. 
A demo interface over the new branch was developed to showcase the system, however to have a complete integration with CDS, a more deep work is required. 
Our system is deployed as a standalone jar which should be incorporated into a Solr instance. Also, some extra configurations should be added. Details on this can be found in appendix \ref{appendix-conf_files.}
The communication between Solr and Invenio was through the library {\codefont solrpy}.

### 2014-02-21 docs/thesis/chap-evaluation.tex
work on document: added graphics

\section{Results quality}
In this section we are interested in evaluating the quality of the results provided by our system.
The type of queries that MASE was designed for, are typical equations that can be found in a scientific publication. 
Out main user case is that a user of CDS is interested in finding documents inside the collection whose equations are related to the one that is provided.

\subsection{Equations}
The evaluated queries were provided by physics experts working at CERN from their own research topics. The queries were provided in \LaTeX format as the users used them in their own publications.
Table \ref{equations_table} presents the set of queries that were evaluated, and a simple explanation about it.

>>>>

\section{Results quality}
In this section we are interested in evaluating the quality of the results provided by our system.
The type of queries that $5e^{x+y}$ was designed for, are typical equations that can be found in a physics scientific publication. 
Out main user case is that a user of CDS is interested in finding documents inside the collection whose equations are related to the one that is provided.

\subsection{Equations}
The evaluated queries were provided by physics experts working at CERN from their own research topics. The queries were provided in \LaTeX format as the users used them in their own publications.
Table \ref{equations_table} presents the set of queries that were evaluated, and a simple explanation about it. 
Each of the users evaluated the results only for the equations he/she provided. For each of the presented result, the user ranked it as "Strictly Relevant", "Partially Relevant" and "Completely Irrelevant". As all evaluations including human interaction, this scoring is at some level subjective.

We present results for three metrics: Partial precision (Document either partially relevant or completely relevant) strict precision and discounted cumulative gain.

### 2014-02-21 docs/thesis/chap-technical_considerations.tex
work on document: added graphics

\section{Digital Representations of Mathematics}
When building any kind of information retrieval system, selecting the right representations of the data can affect the type of algorithms that one can use, the amount of storage, the quality of the data among other factors. Mathematical content is not an exception, and there are available different formats to encode a mathematical expression. One of the most common ways to represent mathematics in scientific documents, which suits CDS's domain, is \LaTeX. Another  important standard for the representation of mathematics is MathML, which was first proposed by the World Wide Web Consortium in 1998 as the recommended language to be used in web documents. One of the advantages of MathML over LaTeX, is the fact that it is an application of XML, which makes it very easy process algorithmically. There two different variations of MathML, presentation and content. The first one is oriented to the visual representation of the expression and the second one is more oriented to the semantics of the expression. OpenMath is another markup language to represent mathematics and can be used to compliment presentation MathML to add more semantic information.

Table \ref{math_languages} presents a comparison on how different languages can express the expression $x^2 + y^2 = z^2$

\begin{longtable}{|m{3.5cm}|>
{\centering\arraybackslash}m{3cm}|>
{\centering\arraybackslash}m{3cm}|>
{\centering\arraybackslash}m{3cm}|
}
\hline 
\textbf{\LaTeX} & 
\textbf{Presentation MathML} & 
\textbf{Content MathML} & 
\textbf{OpenMath}  \\ \hline

\verb|$x^2 + y^2 = z^2$|&&&
\\
\hline

@@

Because

>>>>

\section{Digital Representations of Mathematics}
When building any kind of information retrieval system, selecting the right representations of the data can affect the type of algorithms that one can use, the amount of storage, the quality of the data among other factors. Mathematical content is not an exception, and there are available different formats to encode a mathematical expression. One of the most common ways to represent mathematics in scientific documents, which suits CDS's domain, is \LaTeX.

Another  important standard for the representation of mathematics is MathML, which was first proposed by the World Wide Web Consortium in 1998 as the recommended language to be used in web documents. One of the advantages of MathML over LaTeX, is the fact that it is an application of XML, which makes it very easy process algorithmically. There two different variations of MathML, presentation and content. The first one is oriented to the visual representation of the expression and the second one is more oriented to the semantics of the expression. Its main component is the <apply> tag which represents the function application .

OpenMath is yet another markup language to represent mathematics and can be used to compliment presentation MathML to add more semantic information. It is mainly used now for defining mathematical concepts into standard content dictionaries which can be imported. MathML is compatible with this dictionaries.

Table \ref{math_languages} presents a comparison on how different languages can express the quadratic equation $x = \frac{-b \pm \sqrt{b^2 - 4ac}}{2a} $

\begin{longtable}{|m{3.5cm}|>
{\centering\arraybackslash}m{3cm}|>
{\centering\arraybackslash}m{3cm}|>
{\centering\arraybackslash}m{5cm}|
}
\hline 
\textbf{\LaTeX} & 
\textbf{Presentation MathML} & 
\textbf{Content MathML + Presentation Markups} & 
\textbf{OpenMath}  \\ \hline

\scriptsize{\lstinline|$x = \frac{-b \pm \sqrt{b^2 - 4ac}}{2a} $|}
& \scriptsize{\codefont <math display="block"> <mrow>   <mi>x</mi>  <mo>=</mo>  <mfrac> <mrow> <mo>-</mo> <mi>b</mi> <mo>
{\unicodefont \char"00B1} </mo> <msqrt> <msup> <mi>b</mi> <mn>2</mn> </msup> <mo>-</mo> <mn>4</mn> <mo></mo> <mi>a</mi> <mo></mo> <mi>c</mi> </msqrt> </mrow> <mrow> <mn>2</mn> <mo></mo> <mi>a</mi> </mrow> </mfrac> </mrow> </math>}
&
\scriptsize{\codefont <math> <apply> <eq/> <ci>x</ci> <csymbol> <mfrac> <apply> <minus/> <mi>b</mi> {\unicodefont \char"00B1} <mroot> <apply> <minus/> <csymbol> <msup> <mi>b</mi> <mn>2</mn> </msup> </csymbol> <csymbol> <mn>4</mn> <mo>\&times;</mo> <mi>a</mi> <mo>\&times;</mo> <mi>c</mi> </csymbol> </apply> <mrow></mrow> </mroot> </apply> <apply> <times/> <cn>2</cn> <ci>a</ci> </apply> </mfrac> </csymbol> </apply> </math> } 
&
\scriptsize{\codefont 
<OMOBJ> <OMA> <OMS cd = "relation1" name="eq"/> <OMV name="x"/>
    <OMA> <OMS cd="arith1" name="divide"/> <OMA> <OMS cd = "arith1" name="minus"/><OMA> <OMS cd = "arith1" name="times"/>
          <OMA> <OMS cd = "arith1" name="times"/> <OMV name="b"/> <OMV name="+ALE-"/> </OMA> <OMA> <OMS cd="arith1" name="root"/> <OMA> <OMS cd = "arith1" name="minus"/> <OMA> <OMS cd="arith1" name="power"/> <OMV name="b"/> <OMI>2</OMI> </OMA> <OMA> <OMS cd = "arith1" name="times"/> <OMA> <OMS cd = "arith1" name="times"/> <OMI>4</OMI> <OMV name="a"/> </OMA> <OMV name="c"/> </OMA> </OMA> <OMA> <OMI> 2 </OMI> </OMA> </OMA> </OMA> <OMA> <OMS cd = "arith1" name="times"/> <OMI>2</OMI> <OMV name="a"/> </OMA> </OMA> </OMA> </OMOBJ>}
\\
\hline

@@

\LaTeX allows the most compact representations of the equations and it is the closest language to the community. However, its processing is not straightforward because a given equation can include multiple commands from different packages. Also there are several ways to represent the same set of symbols, either using its Unicode code-point or through embedded commands. Another difficulty are custom commands that users can define in the preamble of the document. In overall, the interpretation and processing of a mathematical expression in this language would require writing a complete compiler which is not the purpose of this work. MathML offers a better compromise between the length of the representation and the easiness to be processed.
